---
title: "Tomey: Map Making"
output: html_notebook
---

```{r}
require(tidyverse)
require(sf)
require(tigris)
require(geodata)
require(terra)
require(colorspace)
require(tidyterra)
require(ggnewscale)
require(cowplot)
```
# Short Explanatory Paragraph

## A couple sentence background about the study and study area you are mapping
Raccoon rabies is distributed along the eastern United States with Alabama serving as the westernmost front of the disease as well as management efforts. However, the fate of Oral Rabies Vaccine baits within Alabama is unknown. This study uses camera traps to evaluate the uptake of baits by target and non-target competitors. 

This map highlights the northern area of the study that falls into the Eastern highland rim ecoregion. Within the region, we conducted sampling at Redstone Arsenal which is characterized by woody wetlands, relatively urban surroundings, and deciduous forest. 

## Information on where you got your spatial elements and why you chose them.

### Vector Elements 
1. Camera Locations
   Actual camera locations were recorded in the field using latitude and longitude. I loaded the CSV containing each camera location for the map. I used the location of cameras rather than the 'site' locations for data collection because it portrays more accurate locational data related to individual cameras related to a polygon overstretching the 12 cameras within each site. 
   
2. Redstone Arsenal Boundary
  I created this polygon feature in ArcGIS pro by manually drawing the shape around the boundary portrayed by baselayer maps because I was unable to find a shapefile to directly download. I then uploaded the shapefile to the map. I chose to use the boundary for Redstone Arsenal rather than creating an extent box around the camera locations because I was held to sampling on the property. 
  
3. Alabama Boundary
  The boundary of Alabama was downloaded from the USA states data available through RStudio. This was downloaded for the addition to my zoomed-out map and shows where RedStone Arsenal is located within the state as well as the relevant size. 

### Raster Element
4. Elevation (30s)
  The elevation within Alabama along with my study area was downloaded directly from a package in RStudio. Camera/site locations in the study were split between upland and lowland areas. Elevation can be a simple but effective way to represent whether cameras represented upland or lowland habitat within the study area without diving into fine scale characteristics involved in defining those habitats. 


# Raster and Vector Elements

## Camera Locations and Study Area Boundary
```{r}
camData = read.csv('https://raw.githubusercontent.com/rytomey/WILD7970_Spatial/main/W2_MapMaking/RSA_Cameras.csv')

camera = st_as_sf(camData, coords = c("Longitude", "Latitude"), crs='+proj=longlat +datum=WGS84 +no_defs +type=crs')

RSA_bound = st_read('/vsicurl/https://github.com/rytomey/WILD7970_Spatial/raw/main/W2_MapMaking/Redstone_Arsenal.shp')

#NAD_1983_CORS96_UTM_Zone_16N
RSA_bound = st_set_crs(RSA_bound, '+proj=utm +zone=16 +ellps=GRS80 +datum=NAD83 +units=m +no_defs')
st_crs(RSA_bound)

plot(st_geometry(RSA_bound))
plot(st_geometry(camera))
```


## Outline of Alabama 
```{r}
alabama = states() %>% 
  filter(NAME=='Alabama')

plot(st_geometry(alabama))

st_crs(alabama)$proj4string

alabama = st_transform(alabama, st_crs(camera))
```


## Elevation: Alabama 
```{r}
usaElev = elevation_30s(country='USA', path=tempdir())

alabama = st_transform(alabama, st_crs(usaElev))
ALelev = crop(usaElev, alabama)

plot(ALelev)
plot(st_geometry(alabama), add=T)

ALelev = mask(ALelev, alabama)
plot(ALelev)
```

# Zoomed-Out Map
## Alabama showing Study Area + Cameras
```{r}
plot(st_geometry(alabama))
plot(RSA_bound, col='blue', add=T)
plot(st_geometry(camera), add=T)
```


## Elevation within Redstone Arsenal 
```{r}
RSA_bound = st_transform(RSA_bound, st_crs(alabama))
RSAelev = crop(ALelev, RSA_bound)
plot(RSAelev, col=sequential_hcl(13, "Terrain"))

RSAelev = mask(RSAelev, RSA_bound)
plot(RSAelev, col=sequential_hcl(13, "Terrain"))

```
# MAPS 
## Zoomed-Out

```{r}
zoomedOut = ggplot()+
  geom_sf(data = alabama)+
  geom_sf(data = RSA_bound, fill=NA, color='Black')

zoomedOut

zoomedOut = zoomedOut +
  theme_bw()+ #Gets rid of most of the colors
  theme(panel.grid=element_blank(), #Removes the gridlines
        axis.text=element_blank(), #Removes the axis text
        axis.ticks=element_blank(), #Removes the axis tick marks
        panel.border=element_blank()) #Removes the frame around the graph

zoomedOut

```

## Zoomed-In
Information on adding a scale bar and north arrow found at: https://stackoverflow.com/questions/61809382/how-can-i-put-a-scalebar-and-a-north-arrow-on-the-map-ggplot

```{r}
zoomedIn = ggplot()+
  geom_spatraster(data = RSAelev)+
  scale_fill_gradient(low='black', high='white', na.value='grey', labs(fill='Elevation'))

zoomedIn = zoomedIn +
  geom_sf(data = camera, aes(pch=Habitat), color='green')+
  scale_shape_manual(values=c(15, 17), name='Camera X Habitat')

zoomedIn = zoomedIn +
  theme_bw()+ #Gets rid of most of the colors
  theme(panel.grid=element_blank(), #Removes the gridlines
        axis.text=element_blank(), #Removes the axis text
        axis.ticks=element_blank(), #Removes the axis tick marks
        panel.border=element_blank()) #Removes the frame around the graph

install.packages(ggspatial)
library(ggspatial)

zoomedIn = zoomedIn +
  ggspatial::annotation_scale(
    location = "tr", 
    bar_cols = c("grey60", "white"),
    text_cex = 0.8) + 
  ggspatial::annotation_north_arrow(
    location = "bl", which_north = "true",
    pad_x = unit(0.01, "in"), pad_y = unit(0.01, "in"),
    style = ggspatial::north_arrow_nautical(
      fill = c("grey40", "white"),
      line_col = "grey20"))

zoomedIn

```

## Cowplotting 

```{r}
plot_grid(zoomedOut, zoomedIn)

legend = get_legend(zoomedIn)
ggdraw(legend)

```





