---
title: "BUILD YOUR OWN MAP"
output: html_document
---
```{r}
require(tidyverse)
require(sf)
require(tigris)
require(geodata)
require(terra)
require(colorspace)
require(tidyterra)
require(ggnewscale)
require(cowplot)

```

## A couple sentence background about the study and study area you are mapping


## Information on where you got your spatial elements and why you chose them


## Raster and vector elements
At least one spatial vector element (created or downloaded) (1 point)
At least one spatial raster element (created or downloaded) (1 point)
A third element (can be vector or raster) (1 point)
Note that in the example we went through in lab, I consider that the final product has 2 raster elements (elevation and old growth cover), and two vector elements (the sampling points, and the state of Oregon's boundary). The extent rectangle doesn't count.
An interpretable legend (1 points)
Fully reproducible (I want to be able to open your .rmd, press run, and have your output)  (1 point)


## Zoomed-out map (4 points)
Extent rectangle on the zoomed-out map (1 point)

Adding the study area and bounding box
```{r}
#Read in study area shapefile 
siteData = st_read()

#Extract x and y coordinates of the bounding box
studyArea = st_bbox(siteSf)
studyArea

#Let's add 2000 m to each dimension
studyArea = studyArea + c(-2000, -2000, 2000, 2000)
studyArea

#And now we convert it to a spatial sfc object
studyArea = st_as_sfc(studyArea)

plot(studyArea)
```

Adding the state boundary and transforming the state boundary to the same coordinate system as the study area 
```{r}
oregon = states() %>% 
  filter(NAME=='Oregon')

plot(st_geometry(oregon))

st_crs(oregon)$proj4string

oregon = st_transform(oregon, st_crs(siteSf))
```

Plotting 
```{r}
plot(st_geometry(oregon))
plot(studyArea, col='red', add=T)
plot(st_geometry(siteSf), add=T)
```


## Zoomed-in map (4 points)


## A short paragraph (within your .rmd file) describing the figure in which you include (5 points)

Do your best to make it relatively neat (1 point)
~ One bonus point if you can figure out how to add a north arrow and scale bar to your figure.
